Gap formation and stability in non-isothermal protoplanetary discs

### 2014-12-22 paper/model.tex
final edits before submission

The initial azimuthal velocity $v_{\phi i}$ is set by centrifugal balance with
pressure, stellar gravity and disc gravity. For a thin disc, 
$v_{\phi}\simeq r\Omega_k$. The initial radial velocity is
$v_{r}=3\nu/r$, where $\nu = \hat{\nu}\rin^2\Omega_k(\rin)$, and we
adopt $\hat{\nu}= 10^{-9}$, so that $|v_{r}/v_{\phi}|\ll1$ and the initial 
flow is effectively only in the azimuthal direction.  With this value 
of physical viscosity, the only source of heating is through
compression, shock-heating (via artificial viscosity) and the
$\mathcal{C}$ function when $e/\Sigma<e_i/\Sigma_i$.

\subsection{Planet potential}\label{planet_config}
The planet potential is given by 
\begin{align}
  \Phi_p = -\frac{GM_p}{\sqrt{|\bm{r} - \bm{r}_p|^2 + \epsilon_p^2}},
\end{align}
where $M_p$ is the planet mass and we fix $q\equiv M_p/M_*=10^{-3}$
throughout this work. This corresponds to a Jupiter-mass planet if $M_*=M_{\sun}$. 
The planet's position in the disc 
$\bm{r}_p=(r_p,\phi_p)$  and $\epsilon_p=0.5r_h$ is a softening
length with $r_h=(q/3)^{1/3}r_p$ being the Hill radius.  
%Hill radius chosen so eps doesn't change with time (scale-height does)
The planet is held on a fixed circular orbit with $ r_p = 10\rin$. This also
defines the time unit $P_0\equiv 2\pi/\Omega_k(r_p)$ used to describe results. 
%In this study only one planet mass is considered, with $q=10^{-3}$. 
%The initial orbital radius is

\section{Numerical experiments}\label{method}
The disc-planet system is evolved using the 
\texttt{FARGO-ADSG} code \citep{baruteau08, baruteau08b}. This is a modified version 
of the original \texttt{FARGO} code \citep{masset00a} to include the energy 
equation. The code employs a finite-difference scheme similar 
to the \texttt{ZEUS} code \citep{stone92}, but with a modified azimuthal transport 
algorithm to circumvent the time-step restriction set by the inner disc boundary. 
The disc is divided into $(N_r,N_\phi)$ zones in the radial and azimuthal directions, 
respectively. The grid spacing is logarithmic in radius and uniform in azimuth.

@@

The generalised PV appears in the description of the linear stability
of radially-structured adiabatic discs \citep{lovelace99,li00}, where
the authors show an extremum in $\teta$ may lead to a dynamical
instability, the RWI. In a barotropic disc, the entropy factor is
absent and the important quantity is the PV.

>>>>

The initial azimuthal velocity $v_{\phi i}$ is set by centrifugal balance with
pressure forces and stellar gravity. For a thin disc, 
$v_{\phi}\simeq r\Omega_k$. The initial radial velocity is
$v_{r}=3\nu/r$, where $\nu = \hat{\nu}\rin^2\Omega_k(\rin)$, and we
adopt $\hat{\nu}= 10^{-9}$, so that $|v_{r}/v_{\phi}|\ll1$ and the initial 
flow is effectively only in the azimuthal direction.  With this value 
of physical viscosity, the only source of heating is through
compression, shock-heating (via artificial viscosity) and the
$\mathcal{C}$ function when $e/\Sigma<e_i/\Sigma_i$.

\subsection{Planet potential}\label{planet_config}
The planet potential is given by 
\begin{align}
  \Phi_p = -\frac{GM_p}{\sqrt{|\bm{r} - \bm{r}_p|^2 + \epsilon_p^2}},
\end{align}
where $M_p$ is the planet mass and we fix $q\equiv M_p/M_*=10^{-3}$
throughout this work. This corresponds to a Jupiter-mass planet if $M_*=M_{\sun}$. 
The planet's position in the disc 
$\bm{r}_p=(r_p,\phi_p)$  and $\epsilon_p=0.5r_h$ is a softening
length with $r_h=(q/3)^{1/3}r_p$ being the Hill radius.  
%Hill radius chosen so eps doesn't change with time (scale-height does)
The planet is held on a fixed circular orbit with $ r_p = 10\rin$ and $\phi_p=\Omega_k(r_p)t$. 
This also
defines the time unit $P_0\equiv 2\pi/\Omega_k(r_p)$ used to describe results. 
%In this study only one planet mass is considered, with $q=10^{-3}$. 
%The initial orbital radius is

\section{Numerical experiments}\label{method}
The disc-planet system is evolved using the 
\texttt{FARGO-ADSG} code \citep{baruteau08, baruteau08b}. This is a modified version 
of the original \texttt{FARGO} code \citep{masset00a} to include the energy 
equation. The code employs a finite-difference scheme similar 
to the \texttt{ZEUS} code \citep{stone92}, but with a modified azimuthal transport 
algorithm to circumvent the time-step restriction set by the fast rotation speed at the 
inner disc boundary. 
The disc is divided into $(N_r,N_\phi)$ zones in the radial and azimuthal directions, 
respectively. The grid spacing is logarithmic in radius and uniform in azimuth.

@@

The generalised PV appears in the description of the linear stability
of radially-structured adiabatic discs \citep{lovelace99,li00}, where
the authors show an extremum in $\teta$ may lead to a dynamical
instability, the RWI. In a barotropic disc where $p=p(\Sigma)$, the entropy factor is 
absent and the important quantity is the PV.

### 2014-12-22 paper/paper.tex
final edits before submission

\maketitle
\begin{abstract}
  Several observations of transition discs show lopsided
  dust-distributions. A potential explanation is the formation of a
  large-scale vortex acting as a dust-trap at the edge of a gap opened
  by a giant planet. Numerical models of gap-edge vortices have
  thus far employed locally isothermal discs, but the 
  theory of this vortex-forming or `Rossby wave' instability was
  originally developed for adiabatic discs.  
  We generalise the study of planetary gap stability to non-isothermal
  discs using customised numerical simulations of disc-planet
  systems where the planet opens an unstable gap. 
  We include in the energy equation a simple cooling function with
  cooling timescale $t_c=\beta\Omega_k^{-1}$, where $\Omega_k$ is
  the Keplerian frequency, and examine the effect of $\beta$ on the
  stability of gap edges and vortex lifetimes. We find increasing
  $\beta$ lowers the growth rate of non-axisymmetric perturbations, and the
  dominant azimuthal wavenumber $m$ decreases.  
  We find a quasi-steady state consisting of one  
  large-scale, over-dense vortex circulating the outer gap edge, typically  
  lasting $O(10^3)$ orbits. 
  Vortex lifetimes were found to generally increase with cooling times up to an
  optimal value, beyond which vortex lifetimes decrease. This
  non-monotonic dependence is qualitatively consistent with 
  recent studies using strictly isothermal discs that vary the disc aspect ratio. 
%  presented in Fu et al. (2014, ApJL, 788, L41). 
\end{abstract}

@@

However, studies of Rossby vortices at planetary gap-edges have 
adopted locally isothermal discs, %cite lin11, zhu14, lyra
where the disc temperature is a fixed function of
position only \citep[e.g.][]{lyra08,lin11a,zhu14,fu14}. 
%isothermal valid for outer disc  
On the other hand, the theory of the RWI was in fact
developed for adiabatic discs \citep{li00}, which permits
heating. %cite 
In adiabatic discs, the relevant quantity for stability
becomes a generalisation of the PV that include entropy variations
\citep{lovelace99}.

>>>>

\maketitle
\begin{abstract}
  Several observations of transition discs show lopsided
  dust-distributions. A potential explanation is the formation of a
  large-scale vortex acting as a dust-trap at the edge of a gap opened
  by a giant planet. Numerical models of gap-edge vortices have
  so far employed locally isothermal discs, but the 
  theory of this vortex-forming or `Rossby wave' instability was
  originally developed for adiabatic discs.  
  We generalise the study of planetary gap stability to non-isothermal
  discs using customised numerical simulations of disc-planet
  systems where the planet opens an unstable gap. 
  We include in the energy equation a simple cooling function with
  cooling timescale $t_c=\beta\Omega_k^{-1}$, where $\Omega_k$ is
  the Keplerian frequency, and examine the effect of $\beta$ on the
  stability of gap edges and vortex lifetimes. We find increasing
  $\beta$ lowers the growth rate of non-axisymmetric perturbations, and the
  dominant azimuthal wavenumber $m$ decreases.  
  We find a quasi-steady state consisting of one  
  large-scale, over-dense vortex circulating the outer gap edge, typically  
  lasting $O(10^3)$ orbits. 
  Vortex lifetimes were found to generally increase with cooling times up to an
  optimal value, beyond which vortex lifetimes decrease. This
  non-monotonic dependence is qualitatively consistent with 
  recent studies using strictly isothermal discs that vary the disc aspect ratio. 
%  presented in Fu et al. (2014, ApJL, 788, L41). 
\end{abstract}

@@

However, studies of Rossby vortices at planetary gap-edges have 
adopted locally isothermal discs, %cite lin11, zhu14, lyra
where the disc temperature is a fixed function of
position only \citep[e.g.][]{lyra08,lin11a,zhu14,fu14}. 
%isothermal valid for outer disc  
On the other hand, the theory of the RWI was in fact
developed for adiabatic discs \citep{li00}, which permits
heating. %cite 
In adiabatic discs, the relevant quantity for stability
becomes a generalisation of the PV that accounts for entropy variations
\citep{lovelace99}.

### 2014-12-22 paper/results.tex
final edits before submission

\section{Growth of non-axisymmetric modes without the influence of the
  planet}\label{linear1} 
In this section, the planet is introduced at $t=20P_0$ and 
its potential is switched on over $10P_0$. At $t=30P_0$ we switch off the
planet potential and azimuthally average the surface density, energy
and velocity fields. (At this point the planet has carved a partial
gap and the RWI has not yet occurred.) Effectively, we initialise the
disc with a gap profile. 
We then perturb the surface density in the outer disc ($r>r_p$) and continue to 
evolve the disc. We impose  sinusoidal perturbations with 
azimuthal wavenumbers $m\in[1,10]$ and random amplitudes within $\pm 0.01$
times the local surface density. 
This procedure allows us to analyse the growth of 
non-axisymmetric modes associated with the gap, but without
complications from non-axisymmetry arising directly from disc-planet
interaction (i.e. planet-induced wakes).

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=1.2cm
  0cm 0cm 0cm]{figures/linear_stability}
  \caption{Evolution of azimuthal Fourier modes of disc surface
    density, non-dimenionlised by the initial axisymmetric component 
    $\Sigma_0(t=0)$, for the
    `planet-off' simulation with $\tilde{\beta}=10$. Colours correspond
    to different $m$ values. The $m=3$ component is the fastest growing
    mode during linear growth with a growth rate of 
    $\gamma=0.017\Omega(r_p)$.\label{linearmodes}}
  % {\bf is the $\Sigma_0$ at t=0?} }  
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
  0cm 0cm 1.1cm]{figures/longterm_planetoff}
  \caption{Long term simulations without the planet potential after
    the gap is set up. The $m=1$ surface density component,
    non-dimenionlised by the initial $m=0$ component, at the
    outer gap edge is shown as a function of the cooling timescale. 
  } \label{planetofflifetimeplot}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
  0cm 0cm 1cm]{figures/longterm_stability}
  \caption{Evolution of the non-dimesnionalized $m=1$ surface density component
    for long term
    simulations with the planet potential kept on. Notice there is
    vortex growth after the initial linear growth, which
    contrasts to Fig. \ref{planetofflifetimeplot}, where vortices
    decay in the absence of continuous disc-planet interaction. 
    \label{lifetimeplot}}   
\end{figure}

In all cases the system remains in a quasi-steady state for
$\gtrsim800P_0$ with a single vortex circulating 
the outer gap edge at the local Keplerian  
frequency. Fig. \ref{Vortex2D} shows a typical plot of the relative 
surface density perturbation in this state. During this stage, the 
vortex intensifies. This is better shown in Fig. \ref{rossbyplot} as the
evolution of Rossby numbers measured at the vortex centres.  
%As the vortex forms it has an initial 
%Rossby number $Ro\approx-0.1$ for all $\tilde\beta$
%(implying anti-cyclonic 
%motion). 
The Rossby number increases in magnitude during quasi-steady state,
but the maximum $|Ro|$ is similar for all $\tilde{\beta}$: the vortex
reaches a characteristic value of $Ro\approx-0.35$ for 
$\tilde\beta=0.1$ and $Ro\approx -0.45$ for $\tilde\beta=10$.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
  0cm 0cm 1cm]{figures/rossby}
  \caption{Evolution of Rossby numbers at the centres of the vortices
    formed in discs with different cooling rates. A negative Rossby
    number implies anti-cyclonic motion. Boxcar averaging
    was used to remove contributions from the planet-induced spiral shock.\label{rossbyplot}}    
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
  0cm 0cm 1cm]{figures/vortex_density}
  \caption{Average value of the relative surface density perturbation
    at vortex centres for various cooling
    times. Intial vortex overdensities decrease with cooling rate while
    vortex growth rates increase with cooling rate.
    \label{overdensity}}     
\end{figure}

Fig. \ref{lifetimeplot} shows that the duration of the quasi-steady state
varies with the cooling rate: for 
$\tilde{\beta}=0.1$ and $10$, the vortex amplitude begins to decay around
$t\sim800P_0$, while for $\tilde{\beta}=1,\,5$ the decay begins at 
$t\sim1200P_0$. This non-monotonic dependence suggests that there
exists an optimal cooling rate to maximise the vortex lifetime. We
will discuss this issue further in 
\S\ref{lifetime_discuss}. Notice the decay timescale decreases with
increasing cooling time: for $\tilde{\beta}=0.1$ it takes $\sim400P_0$
whereas for $\tilde{\beta}=10$ it takes $\sim 100P_0$ for the $m=1$
amplitude to decay significantly after reaching maximum.

@@

\begin{figure}
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock1}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock2}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock3}
  } \\[-0.98cm]
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock4}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock5}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock6}
  } 
  \caption{The vortex in the case with $\tilde{\beta}=1$
    during quasi-steady state (left), start of decay 
    (middle), and just after the decay in the $m=1$ amplitude
    (right). The surface density perturbation
    (top) and the associated surface density gradient (bottom) are
    shown. Wake like features corresponding to large density gradients
    are found to originate from the vortices during the late phase of 
    their quasi-steady states and into dissipation times.
    This plot is to be considered in conjunction with
    Fig. \ref{lifetimeplot}. 
    \label{shockplot}}
  % Large surface density gradients become prominant
  % around vortex during dissipation} 
\end{figure}

@@

Putting the above observations together, we suggest that vortex decay
(in the $m=1$ surface density amplitude) is due to shock formation by
the vortex. When the vortex reaches large amplitude, it begins
to induce shocks in the surrounding fluid, as supported by the
increase in Mach number and the appearance of wakes with large surface
density gradients. The vortex may lose energy through shock
dissipation. In addition, a strong vortex (or shock formation) 
can smooth out the gap structure that originally gave rise to the
RWI, which would oppose vortex growth. We examine this below.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/gapchange}
  \caption{Azimuthally averaged profiles of the relative surface
    density perturbations for  $\tilde\beta=1.0$
    before (solid) and after vortex decay
    (dashed). Alongside drastic reduction in the outer gap maxima, vortex decay
    smoothens out the sharp outer gap edge.
    \label{gap_smoothed}} 
\end{figure}

@@

A plot of the gradient parameter over time
for the $\tilde\beta=1.0$ case is shown in Fig. \ref{smoothnessplot}.
During vortex decay, the outer gap edge is 
drastically smoothed out, changing from a value of $\delta\Sigma=1.2$ during
quasi-steady state to $0.4$ after dissipation. 
%and we find that the gap edge is smoothed during this
%process. 
This can be interpreted as the vortex providing a viscosity;  
and we measure a typical alpha viscosity $\alpha = O(10^{-2})$
associated with the vortex. This acts against gap-opening
by the planet, so the condition for the RWI
becomes less favourable, which may explain why vortices do not
reform again (at least within the simulation timescale). 
%Thus, the
%time needed for the vortex to grow to sufficient amplitude to modify
%the gap structure

\begin{figure}
  \includegraphics[width=\linewidth]{figures/gap_smoothness}
  \caption{Non-dimensional measure of the surface density gradient at
    the outer gap edge $\tilde\beta=1.0$. During the vortex quasi-steady state
    the gap edge is found to have a large gradient and sharp peak while vortex
    dissipation,
    which occurs at $t\approx1000P_0$ as seen in Fig. \ref{lifetimeplot},
    works to smoothen out the gap edge.
% A large $\delta\Sigma$
 %   corresponds to a more sharply peaked outer gap edge.
    \label{smoothnessplot}}  
\end{figure}

>>>>

\section{Growth of non-axisymmetric modes without the influence of the
  planet}\label{linear1} 
In this section, the planet is introduced at $t=20P_0$ and 
its potential is switched on over $10P_0$. At $t=30P_0$ we switch off the
planet potential and azimuthally average the surface density, energy
and velocity fields. At this point the planet has carved a partial
gap but the RWI has not yet occurred. 
We then perturb the surface density in the outer disc ($r>r_p$) and continue to 
evolve the disc. (Effectively, we initialise the
disc with a gap profile.) We impose  sinusoidal perturbations with 
azimuthal wavenumbers $m\in[1,10]$ and random amplitudes within $\pm 0.01$
times the local surface density. 
This procedure allows us to analyse the growth of 
non-axisymmetric modes associated with the gap, but without
complications from non-axisymmetry arising directly from disc-planet
interaction (i.e. planet-induced wakes).

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=1.2cm
  0cm 0cm 0cm]{figures/linear_stability}
  \caption{Evolution of azimuthal Fourier modes of disc surface
    density, non-dimensionalised by the initial axisymmetric component 
    $\Sigma_0(t=0)$, for the
    `planet-off' simulation with $\tilde{\beta}=10$. Colours correspond
    to different $m$ values. The $m=3$ component is the fastest growing
    mode during linear growth with a growth rate of 
    $\gamma=0.017\Omega(r_p)$.\label{linearmodes}}
  % {\bf is the $\Sigma_0$ at t=0?} }  
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
  0cm 0cm 1.1cm]{figures/longterm_planetoff}
  \caption{Long term simulations without the planet potential after
    the gap is set up. The $m=1$ surface density component,
    non-dimensionalised by the initial $m=0$ component, at the
    outer gap edge is shown as a function of the cooling timescale. 
  } \label{planetofflifetimeplot}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
  0cm 0cm 1cm]{figures/longterm_stability}
  \caption{Evolution of the non-dimensionalised $m=1$ surface density component
    for long term
    simulations with the planet potential kept on. Notice there is
    vortex growth after the initial linear growth, which
    contrasts to Fig. \ref{planetofflifetimeplot}, where vortices
    decay in the absence of continuous disc-planet interaction. 
    \label{lifetimeplot}}   
\end{figure}

In all cases the system remains in a quasi-steady state for
$\gtrsim800P_0$ with a single vortex circulating the outer gap edge at
the local Keplerian frequency. Fig. \ref{Vortex2D} shows a typical
plot of the relative surface density perturbation in this state.    
The duration of the quasi-steady state varies with the cooling
rate. Fig. \ref{lifetimeplot} shows that for $\tilde{\beta}=0.1$ and
$10$, the vortex amplitude begins to decay around 
$t\sim800P_0$, while for $\tilde{\beta}=1,\,5$ the decay begins at 
$t\sim1200P_0$. This non-monotonic dependence suggests that there
exists an optimal cooling rate to maximise the vortex lifetime. We
will discuss this issue further in 
\S\ref{lifetime_discuss}. Notice the decay timescale decreases with
increasing cooling time: for $\tilde{\beta}=0.1$ it takes $\sim400P_0$
whereas for $\tilde{\beta}=10$ it takes $\sim 100P_0$ for the $m=1$
amplitude to decay significantly after reaching maximum.

@@

We find the vortices intensify during the quasi-steady state. This is
better shown in Fig. \ref{rossbyplot} as the evolution of Rossby
numbers measured at the vortex centres.   
%As the vortex forms it has an initial 
%Rossby number $Ro\approx-0.1$ for all $\tilde\beta$
%(implying anti-cyclonic 
%motion). 
The Rossby number increases in magnitude during quasi-steady state,
but the maximum $|Ro|$ is similar for all $\tilde{\beta}$: the vortex
reaches a characteristic value of $Ro\approx-0.35$ for 
$\tilde\beta=0.1$ and $Ro\approx -0.45$ for $\tilde\beta=10$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
  0cm 0cm 1cm]{figures/rossby}
  \caption{Evolution of Rossby numbers at the centres of the vortices
    formed in discs with different cooling rates. A negative Rossby
    number implies anti-cyclonic motion.%  Boxcar averaging
    % was used to remove contributions from the planet-induced spiral shock.
    \label{rossbyplot}}    
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
  0cm 0cm 1cm]{figures/vortex_density}
  \caption{Evolution of the average value of the relative surface density perturbation
    at vortex centres for various cooling
    times. % The initial vortex over-densities decrease with increasing cooling times while
    % vortex growth rates increase with cooling rate.
    \label{overdensity}}     
\end{figure}

@@

\begin{figure}
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock1}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock2}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock3}
  } \\[-0.98cm]
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock4}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock5}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock6}
  } 
  \caption{The vortex in the case with $\tilde{\beta}=1$
    during quasi-steady state (left), start of decay 
    (middle), and just after the decay in the $m=1$ amplitude
    (right). The surface density perturbation
    (top) and the associated surface density gradient (bottom) are
    shown. Wakes corresponding to large density gradients
    are found to originate from the vortices during the late phase of 
    their quasi-steady states, and persists during their decay. 
    This plot is to be considered in conjunction with
    Fig. \ref{lifetimeplot}. 
    \label{shockplot}}
  % Large surface density gradients become prominant
  % around vortex during dissipation} 
\end{figure}

@@

Putting the above observations together, we suggest that vortex decay
(in the $m=1$ surface density amplitude) is associated with shock formation by
the vortex. When the vortex reaches large amplitude, it begins
to induce shocks in the surrounding fluid, as supported by the
increase in Mach number and the appearance of wakes with large surface
density gradients. The vortex may lose energy through shock
dissipation. In addition, a strong vortex (or shock formation) 
can smooth out the gap structure that originally gave rise to the
RWI, which is expected to oppose vortex growth. We examine this below.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/gapchange}
  \caption{Azimuthally averaged profiles of the relative surface
    density perturbations for  $\tilde\beta=1.0$
    before (solid) and after vortex decay
    (dashed). Alongside drastic reduction in the outer gap maxima, vortex decay
    smooth out out the sharp outer gap edge.
    \label{gap_smoothed}} 
\end{figure}

@@

A plot of the gradient parameter over time
for the $\tilde\beta=1.0$ case is shown in Fig. \ref{smoothnessplot}.
During vortex decay, the outer gap edge is 
drastically smoothed out, changing from a value of $\delta\Sigma=1.2$ during
quasi-steady state to $0.4$ after dissipation. 
%and we find that the gap edge is smoothed during this
%process. 
This may be interpreted as the vortex providing a viscosity;  
and we measure a typical alpha viscosity $\alpha = O(10^{-2})$
associated with the vortex. This acts against gap-opening
by the planet, so the condition for the RWI
becomes less favourable, which may explain why vortices do not
reform again (at least within the simulation timescale). 
%Thus, the
%time needed for the vortex to grow to sufficient amplitude to modify
%the gap structure

\begin{figure}
  \includegraphics[width=\linewidth]{figures/gap_smoothness}
  \caption{Non-dimensional measure of the surface density gradient at
    the outer gap edge $\tilde\beta=1.0$. During the vortex quasi-steady state
    the gap edge is found to have a large gradient and sharp peak while vortex
    dissipation,
    which occurs at $t\approx1000P_0$ as seen in Fig. \ref{lifetimeplot},
    works to smooth out out the gap edge.
% A large $\delta\Sigma$
 %   corresponds to a more sharply peaked outer gap edge.
    \label{smoothnessplot}}  
\end{figure}

### 2014-12-22 paper/summary.tex
final edits before submission

We considered two types of numerical experiments. We first used
disc-planet interaction to self-consistently set up gap profiles,
which were then perturbed and evolved without further the influence of
the planet potential. This procedure isolates the effect of cooling on
gap stability through the set up of the initial gap profile. We find
that as the cooling time $t_c$ is increased, the gaps became more
stable, with lower growth rates of non-axisymmetric modes and the
dominant azimuthal wavenumber also decreases. This is consistent with
the notion that increasing $t_c$ leads to higher temperatures or $h$,
which opposes gap-opening by the planet. This means that the gaps
opened by the planet in a disc with longer $t_c$ are smoother and
therefore  more stable to the RWI. %generalized vortensity minima less
                                %pronounced

@@

We remark that a non-monotonic dependence of the vortex lifetime was
also reported by \cite{fu14}, who performed isothermal disc-planet
simulations with different values of the   
disc aspect ratio. In their simulations the optimum aspect ratio is
$h=0.06$. In our simulations, the disc aspect ratio is a dynamical
variable, but by analysing the region where the vortex is located
($r-r_p\in[2,10]r_h$), % {\bf define this region}
we find for $\tilde\beta=5.0$, which has the longest lifetime, that 
$h\approx0.06$ on average.  
%{\bf if new definition of vortex lifetime gives other $\tilde{\beta}$ values as the maximum lifetime, then need to update this value of $h$}. 
Our result is therefore consistent with \citep{fu14}.

@@

\emph{Self-gravity.} We have ignored disc self-gravity in this
study. Based on linear calculations, \cite{lovelace13} concluded
self-gravity to be important for the RWI when the Toomre parameter $Q<O(1/h)$, or
$Q\lesssim 20$ for $h\sim0.05$, as was typically considered in this
work. This suggests that self-gravity may affect vortex lifetimes even
when the $Q$ is not small. In particular, given that
we observe vortices can reach significant over-densities (up to almost
an order of magnitude), it will be important to include disc
self-gravity in the future. %self-gravitational collapse of vortices

>>>>

We considered two types of numerical experiments. We first used
disc-planet interaction to self-consistently set up gap profiles,
which were then perturbed and evolved without further the influence of
the planet potential. This procedure isolates the effect of cooling on
gap stability through the set up of the initial gap profile. We find
that as the cooling time $t_c$ is increased, the gaps became more
stable, with lower growth rates of non-axisymmetric modes and the
dominant azimuthal wavenumber also decreases. This is consistent with
the notion that increasing $t_c$ leads to higher temperatures or
equivalently the disc aspect-ratio $h$,
which opposes gap-opening by the planet. This means that the gaps
opened by the planet in a disc with longer $t_c$ are smoother and
therefore  more stable to the RWI. %generalized vortensity minima less
                                %pronounced

@@

We remark that a non-monotonic dependence of the vortex lifetime was
also reported by \cite{fu14}, who performed isothermal disc-planet
simulations with different values of the   
disc aspect-ratio. In their simulations the optimum aspect-ratio is
$h=0.06$. In our simulations, $h$ is a dynamical
variable, but by analysing the region where the vortex is located
($r-r_p\in[2,10]r_h$), % {\bf define this region}
we find for a dimensionless cooing timescale of $\tilde\beta=5.0$, which has
the longest vortex lifetime, that  
$h\approx0.06$ on average.  
%{\bf if new definition of vortex lifetime gives other $\tilde{\beta}$ values as the maximum lifetime, then need to update this value of $h$}. 
Our result is therefore consistent with \citep{fu14}.

@@

\emph{Self-gravity.} We have ignored disc self-gravity in this
study. Based on linear calculations, \cite{lovelace13} concluded
self-gravity to be important for the RWI when the Toomre parameter $Q<O(1/h)$, or
$Q\lesssim 20$ for $h\sim0.05$, as was typically considered in this
work. This suggests that self-gravity may affect vortex lifetimes even
when $Q$ is not small. In particular, given that
we observe vortices can reach significant over-densities (up to almost
an order of magnitude), it will be important to include disc
self-gravity in the future. %self-gravitational collapse of vortices
